---
title: "DFT test"
author: "Ioann Stanković"
date: "2023-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Creating the first proposed [DFT](https://arxiv.org/pdf/1701.01960.pdf) algorithm, and trying to determine if the binary sequence is truly random.

## Discrete Fourier Transform Test

procedure of the proposed DFT test:

1)

- We generate a random binary sequence with the built in binomial distribution function.

```{r}
set.seed(11)

k = 100

E <- rbinom(k, size = 1, prob = 0.5)

print(E[1:20])
```

- We convert the values in sequence E to -1 and +1 and we create sequence X. 

```{r}
X <- E*2-1

print(X[1:20])
```

2)

- We apply a discrete Fourier transform (DFT) to \(X\) to produce Fourier coefficients \( \{ S_j (X) \}^{n-1}_{j=0} \). The Fourier coefficients \( S_j (X) \) are defined as follows:

\[ S_j (X) = \sum^{n-1}_{k=0}x_k \cos(\frac{2\pi k j}{n}) - \sqrt{-1}\sum^{n-1}_{k=0}x_k\sin(\frac{2\pi k j}{n}), \] where its real and imaginary parts \( c_k \) and \( s_k \) are:

\[ c_k = \sum^{n-1}_{k=0}x_k \cos(\frac{2\pi k j}{n}) \]

\[ s_k = \sum^{n-1}_{k=0}x_k\sin(\frac{2\pi k j}{n}) \]

```{r}
v <- 0:(k-1)

j <- 0:(k-1)

c_values <- numeric(k)
s_values <- numeric(k)

for (j in 0:(k-1)) {
  c_values[j + 1] <- sum(cos(2 * pi * v * j / k) * E)
}

print(c_values[1:10])

for (j in 0:(k-1)) {
  s_values[j + 1] <- sum(sin(2 * pi * v * j / k) * E)
}

print(s_values[1:10])

S <- c_values-sqrt(-1 + 0i)*s_values

print(S[1:10])
```

3) 

- We compute \(\{|S_j(X)|\}^{\frac{n}{2}-1}_{j=0}\), where \[ |S_j(X)|^2 = (c_j(X))^2 + (s_j(X))^2 .\]

Because \(|S_j(X)| = |S_{n-j}(X)|, \) \(\{|S_j(X)|\}^{n-1}_{j=\frac{n}{2}}\)

```{r}

S_abs <- sqrt(c_values^2+s_values^2)

S_disc <- S_abs[1:(k/2-1)]

```

4) This part deviates from the changed algorithm

- We compute the threshold value \( T_{0.95} = \sqrt{3n}. \) The 95% values \(\{|S_j(X)|\}^{\frac{n}{2}-1}_{j=0}\) are supposed to be \( < T_{0.95} \). Lets mention that the value \( \sqrt{3n} \) was reported as an incorrect value by several researchers, so lets use the value \( \sqrt{-n\ln{0.05}} \) as it is more accurate.

```{r}

T <- sqrt(-k*log(0.05))

```

5)

- We count the value \[ N_1 = \# \{ |S_j(X)| \ | \ |S_j(X)< T_{0.95}, 0 \leq j \leq \frac{n}{2}-1  \}.  \]

N_1 can be considered to follow binomial distribution \( B(\frac{n}{2},0.95) \). So according to central limit theorem, when \( n \) is sufficiently large, the approximation to \( B(n,p) \) is given by the normal distribution \( N(np, np(1-p)) \). Therefore, when \( n \) is sufficiently large, under the assumption of randomness,

\[ N_1 ~ N(0.95 \frac{n}{2}, (0.95)(0.05)\frac{n}{2}. \]

```{r}

N <- sum(S_disc < T)

```

6)

- We copmute a test statistic

\[ d = \frac{N_1 - 0.95\frac{n}{2}}{\sqrt{(0.95)(0.05)\frac{n}{2}}}. \]

When \(n\) is sufficiently large, under the assumption of randomness, the test statistic \(d\) can be considered to follow \(N(0,1)\).

```{r}

d <- (N- 0.95* k/2)/(sqrt(0.95*0.05*k/2))

```

7)

- We compute the P-value : \(p = erfc(\frac{|d|}{\sqrt{2}}). \)

If \(p < \alpha \), then we conclude that the sequence is non-random, where \(\alpha \) is a significance level of the DFT test. NIST recommends \(\alpha = 0.01 \). 

If \(p \geq \alpha \), we conclude that the sequence is random.

```{r, warning=FALSE}
library(pracma) #We have to install package (install.packages("pracma"))

p <- erfc(d / sqrt(2))

alpha <- 0.01

p >= alpha

```

8) 

- We perform 1) to 7) for \( m \) sample sequences \( \{ X_1,X_2, \dots, X_m \} \); \( m \) P-values \( \{ p_1, p_2, \dots, p_m \} \)

8): 


We create a list of sequences
```{r}
set.seed(11) #So we can compare the first element of the list with the previous example

k <- 100

m <- 10

E_m <- list()

for (i in 1:m) {
  E_m[[i]] <- rbinom(k, size = 1, prob = 0.5)
}


```

1) -1/+1 transformation

```{r}

X_m <- list()

for (i in seq_along(E_m)) {
  X_m[[i]] <- 2 * E_m[[i]] - 1
}

```

2) Applying DFT to \( X_j,\)  where \( j=1,2, \dots, m \) 

```{r}
c_values_m <- vector("list", length = m)
s_values_m <- vector("list", length = m)

for (i in seq_along(E_m)) {
  c_values_m[[i]] <- numeric(k)  
  s_values_m[[i]] <- numeric(k)  

  for (j in 0:(k-1)) {
    c_values_m[[i]][j + 1] <- sum(cos(2 * pi * v * j / k) * E_m[[i]])
    s_values_m[[i]][j + 1] <- sum(sin(2 * pi * v * j / k) * E_m[[i]])
  }
}


```

3) We compute \( \{ | S_j(X) | \}^{\frac{n}{2}-1}_{j = 0} \)

```{r}

S_abs_m <- vector("list", length = m)

S_abs_m <- Map(function(c, s) sqrt(c^2 + s^2), c_values_m, s_values_m)

S_disc_m <- lapply(S_abs_m, function(x) x[1:(k/2 - 1)])

```

4) We compute the value \( T. \)

```{r}

T <- sqrt(-k*log(0.05))

```

5) We count \( N_m \)

```{r}

N_m <- lapply(S_disc_m, function(x) sum(x < T))

```

6) We compute a test statistic

```{r}

d_m <- lapply(N_m, function(N) (N - 0.95 * k / 2) / sqrt(0.95 * 0.05 * k / 2))

```

7) We compute \(m \) P-values:

```{r}

p_m <- lapply(d_m, function(inner_list) {
  lapply(inner_list, function(x) erfc(x / sqrt(2)))
})

alpha <- 0.01

m_p <- sum(unlist(p_m) >= alpha)
```

9) Second-level test A): Proportion of sequences passing a test

- We count the number of sample sequences for which P-value \( \geq \alpha \) and define it as \( m_p \). Then, under the assumption of randomness, \( m_p \) follows \( B(m, 1-\alpha) \), which approximates \( N(m(1-\alpha),m\alpha(1-\alpha)) \) when \( m \) is sufficiently large. Therefore, the proportion of sequences passing a test \( (=\frac{m_p}{m}) \) approximately follows \( N((1-\alpha), \frac{\alpha(1-\alpha)}{m}) \). The range of acceptable \( \frac{m_p}{m} \) is determined with the significance interval:

\[ 1-\alpha - 3 \sqrt{\frac{\alpha(1-\alpha)}{m}} < \frac{m_p}{m} < 1-\alpha + 3 \sqrt{\frac{\alpha(1-\alpha)}{m}} .\]

If the proportion falls outsidee of this interval, there is evidence that the data are non-random.

```{r}
L <- 1 - alpha - 3* sqrt(alpha*(1-alpha)/m)

R <- 1 - alpha + 3* sqrt(alpha*(1-alpha)/m)

L <= m_p/m & m_p/m <= R


```


10) Second-level test B): Uniform distribution of P-values

## DFT

Lastly creating a function that directly returns \(m\) P-values.

The function takes a binary string of length \( mn \) and transforms it to \(n\) strings with \(m\) values. Returning TRUE/FALSE if the \( \frac{m_p}{m} \) falls in the above given confidence interval. 

```{r, warning=FALSE}
library(pracma)

string <- rbinom(1000, size = 1, prob = 0.5)

n <- 100

m <- 10

DFT <- function(string, n, m) {
  if (n * m != length(string)) {
    string = string[1:n*m]
  }
  E_m <- split(string, ceiling(seq_along(string) / n))
  
  
  X_m <- list()

  for (i in seq_along(E_m)) {
  X_m[[i]] <- 2 * E_m[[i]] - 1
  }
  c_values_m <- vector("list", length = m)
  s_values_m <- vector("list", length = m)
  
  v <- 0:(m-1)

  for (i in seq_along(E_m)) {
    c_values_m[[i]] <- numeric(m)  
    s_values_m[[i]] <- numeric(m)  
    
  
  
    for (j in 0:(m-1)) {
      c_values_m[[i]][j + 1] <- sum(cos(2 * pi * v * j / m) * E_m[[i]])
      s_values_m[[i]][j + 1] <- sum(sin(2 * pi * v * j / m) * E_m[[i]])
    }
  }
  S_abs_m <- vector("list", length = m)

  S_abs_m <- Map(function(c, s) sqrt(c^2 + s^2), c_values_m, s_values_m)

  S_disc_m <- lapply(S_abs_m, function(x) x[1:(m/2 - 1)])

  T <- sqrt(-n*m*log(0.05))
  
  N_m <- lapply(S_disc_m, function(x) sum(x < T))
  
  d_m <- lapply(N_m, function(N) (N - 0.95 * m / 2) / sqrt(0.95 * 0.05 * m / 2))
  
  p_m <- lapply(d_m, function(inner_list) {
    lapply(inner_list, function(x) erfc(x / sqrt(2)))
  })

  alpha <- 0.01

  m_p <- sum(unlist(p_m) >= alpha)
  
  L <- 1 - alpha - 3* sqrt(alpha*(1-alpha)/m)

  R <- 1 - alpha + 3* sqrt(alpha*(1-alpha)/m)

  return(L <= m_p/m & m_p/m <= R)
}


```


